Fast, Robust Classification with Applications to Periodic Variable Stars

### 2012-05-01 report/ms.tex
added a bunch to ms

Periodic variables may be divided into a few dozen classes based on physical properties of light sources. Separating the sources into classes is a critical step in turning raw astronomical observations into scientific knowledge. The size of modern data sets requires that much of this work be automated by machine learning and statistical classifiers.

Figure \ref{fig:cepheid} displays the light curve (i.e. time series) of a periodic variable belonging to the class Classical Cepheid. The points in plot a) represent flux measurements in magnitudes (i.e. brightness of the source) made by the telescope at particular times. The 0 point on the time axis is arbitrary. Using fourier methods, one can estimate a period from the measurements in plot a).  Plot b) of Figure \ref{fig:cepheid} displays the flux measurements of the same object. However here the x-axis is phase of each time measurement, computed using the estimated period of 4.51 days. Here we can observe the structure of the periodic variation. This is known as the \textit{folded light curve}.
\begin{figure}[h]
  \begin{center}
    \begin{includegraphics}[scale=.5]{2000.pdf}
      \caption{Light curve of a Classical type variable star.\label{fig:cepheid}}
    \end{includegraphics}
  \end{center}
\end{figure}

@@

In order to construct a classifier, features are \textit{extracted} from $D$ (i.e. take functions of $D$) that will separate sources into different classes. Features vary from study to study, but typical ones include period (inverse of strongest fourier frequency), amplitude, skew, and estimates of derivatives.

@@

\subsection{Uncertain Features}
A major problem with the standard approach is the high levels and heteroskedastic nature of the uncertainty in the features due to having poorly sampled time series ($l$ is small) or high error in the flux measurements ($e_{i}$ are systematically large). This leads to situations where features meant to represent physical quantities of the time series, such as amplitude or period, are incorrect. As an example \textbf{TODO: Include example with nice graphic here}

@@

>>>>

Periodic variables may be divided into a few dozen classes based on physical properties of light sources. Separating the sources into classes is a critical step in turning raw astronomical observations into scientific knowledge. The size of modern data sets requires that much of this work be automated by machine learning and statistical classifiers. 
\begin{figure}[h]
  \begin{center}
    \begin{includegraphics}[scale=.5]{2000.pdf}
      \caption{Light curve of a Classical Cepheid type variable star. The brightness of the star is on the y-axis. For the top plot the brightness is plotted against time. This time series is periodic. Using fourier methods we can estimate a period of 4.51 days. We can convert the times from the top plot into phase ((time modulo period) / period)). Brightness versus phase is plotted in the bottom plot. Here we can observe the structure in the time series, which usually similar for stars of the same class but quite different for stars of different classes.\label{fig:cepheid}}
    \end{includegraphics}
  \end{center}
\end{figure}

Figure \ref{fig:cepheid} displays the light curve (i.e. time series) of a periodic variable belonging to the class Classical Cepheid. The points in the top plot represent flux measurements in magnitudes (i.e. brightness of the source) made by the telescope at particular times. The 0 point on the time axis is arbitrary. Using fourier methods, one can estimate a period using these measurements.  The lower plot of Figure \ref{fig:cepheid} displays the flux measurements of the same object. However here the x-axis is phase of each time measurement, computed using the estimated period of 4.51 days. Here we can observe the structure of the periodic variation. This is known as the \textit{folded light curve}.

@@

In order to construct a classifier, features are \textit{extracted} from $D_j$ (i.e. take functions of $D_j$) that will separate sources into different classes. Features vary from study to study, but typical ones include period (inverse of strongest fourier frequency), amplitude, skew, and estimates of derivatives.

@@

\subsection{Uncertain Features}
\begin{figure}[h]
  \begin{center}
    \begin{includegraphics}[scale=.5]{period_amplitude.pdf}
      \caption{Period-Amplitude relationship for two classes: \textit{Classical Cepheid} stars and \textit{Population II Cepheid} stars. The green box represents (roughly) the physically possible range of periods and amplitudes. Clearly some of these features have been estimated incorrectly.\label{fig:period_amplitude}}
    \end{includegraphics}
  \end{center}
\end{figure}
A major problem with the standard approach is the high levels and heteroskedastic nature of the uncertainty in the features due to having poorly sampled time series ($l$ is small) or high error in the flux measurements ($e_{i}$ are systematically large). This leads to situations where features meant to represent physical quantities of the time series, such as amplitude or period, are incorrect. As an example see Figure \ref{fig:period_amplitude}. This plot show the period and amplitude features for two classes of stars with a box around known physical limits for these features. Some of the observations have values outside these physical limits.

Many of the features used for classification rely on a correct estimate of period for the object. For example, features related to the derivative are often computed based on the folded light curve. If the light curve is folded on the wrong period, then the derivative estimates are likely to be wrong.

\section{Interval SVMs}
\subsection{Constructing the Intervals from the Time Series}
In certain applications, methods for constructing hyper-rectangles (or hyper-spheres) in feature space may be fairly straitforward. For example in El Ghaoui \cite{el2003robust}, the authors discuss applications to micro-array data where several replicates of some experiment are available, and a hyper-rectangle can be chosen to enclose all replicates.

With astronomy data there is no clearly correct way to construct intervals around features. A few possibilities include:
\begin{enumerate}
\item Make width of feature interval proportional to number of measurements in time series. Long time series mean that features are likely to be accurate. For simple features, such as the standard deviation of the brightness measurements, making the interval width go down at rate $\sqrt{n}$ is grounded in probability theory.
\item Subsample the time series and derive features for each subsample. Represent observations as intervals that contain features derived from every subsample.
\item Heuristically put intervals around observations with features that are wrong. For example the amplitude of a star cannot be greater than 6 magnitudes, so any star with amplitude greater than 6, could be given an interval that contains values less than 6.
\end{enumerate}

- probably happen with any high dim problem, like image classification

\begin{enumerate}
\item sliced each light curve into 5 contiguous parts. ordering measurements in time: slice 1 is 0\% through 50\%, slice 2 is 10\% through 60\%, . . .
\item derive features for each of these parts
\item put the interval minimum at the lowest value of the feature for the 5 slices. put the interval max at the maximum feature value across the 5 slices
\end{enumerate}

@@

\begin{enumerate}
\item problem with normalizing data
\item how is data normalized in svms
\end{enumerate}
